---
title: "HochSchulz 2022 Melanoma selected motif tissue plots"
output:
  html_document:
    code_folding: hide
  md_document:
    variant: gfm
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  fig.width = 12,
  fig.height = 15,
  dpi = 150
)

repo_candidates <- unique(normalizePath(c(
  getwd(),
  file.path(getwd(), ".."),
  file.path(getwd(), "..", ".."),
  file.path(getwd(), "..", "..", "..")
), mustWork = FALSE))
repo_hit <- vapply(repo_candidates, function(path) {
  desc <- file.path(path, "DESCRIPTION")
  file.exists(desc) && any(grepl("^Package: CellEdgeR$", readLines(desc)))
}, logical(1))
root_dir <- if (any(repo_hit)) repo_candidates[which(repo_hit)[1]] else NA_character_
use_local <- !is.na(root_dir)

local_rlib <- if (use_local) file.path(root_dir, ".Rlib") else ".Rlib"
if (dir.exists(local_rlib)) {
  .libPaths(c(normalizePath(local_rlib), .libPaths()))
}

report_dir <- if (use_local) {
  file.path(root_dir, "analysis", "hochchulz_2022_melanoma")
} else {
  normalizePath(getwd(), mustWork = FALSE)
}
```

```{r libs}
if (use_local && requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(root_dir)
} else if (!requireNamespace("CellEdgeR", quietly = TRUE)) {
  stop("Install CellEdgeR or run this report inside the repo with devtools::load_all().")
}

library(CellEdgeR)
library(dplyr)
library(ggplot2)
library(grid)

helpers_candidates <- c(
  if (use_local) file.path(root_dir, "analysis", "hochchulz_2022_melanoma", "hochschulz_2022_melanoma_shared.R"),
  "analysis/hochchulz_2022_melanoma/hochschulz_2022_melanoma_shared.R",
  "hochschulz_2022_melanoma_shared.R"
)
helpers_candidates <- unique(helpers_candidates[!is.na(helpers_candidates)])
helpers_hits <- helpers_candidates[file.exists(helpers_candidates)]
if (!length(helpers_hits)) {
  stop("Missing helper script. Tried: ", paste(helpers_candidates, collapse = ", "))
}
source(helpers_hits[1])
```

```{r params}
n_top_nominal_motifs <- 1
n_per_group <- 6
group_col <- "patient_treatment_group_before_surgery"
group_order <- c("ICI", "untreated")
offset_modes <- "volume"
force_render_panels <- FALSE
offset_mode_labels <- c(
  volume = "volume offset"
)
panel_dir <- file.path(report_dir, "hochschulz_2022_melanoma_selected_motifs_panels")
boxplot_logged_dir <- file.path(report_dir, "hochschulz_2022_melanoma_selected_motifs_boxplots_logged")
if (!dir.exists(panel_dir)) dir.create(panel_dir, recursive = TRUE, showWarnings = FALSE)
if (!dir.exists(boxplot_logged_dir)) dir.create(boxplot_logged_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r cache}
cache_candidates <- unique(c(
  if (use_local) file.path(root_dir, "analysis", "cache", "hochschulz_base.rds"),
  if (use_local) file.path(root_dir, "analysis", "hochchulz_2022_melanoma", "cache", "hochschulz_base.rds"),
  "analysis/cache/hochschulz_base.rds",
  "analysis/hochchulz_2022_melanoma/cache/hochschulz_base.rds"
))
cache_hits <- cache_candidates[file.exists(cache_candidates)]
if (!length(cache_hits)) {
  stop("Base cache not found. Tried: ", paste(cache_candidates, collapse = ", "))
}
base_cache <- cache_hits[1]
base_payload <- readRDS(base_cache)

graphs <- base_payload$graphs
counts <- base_payload$counts
edge_cut <- base_payload$edge_cut
sample_df <- base_payload$sample_df
if (!"sample" %in% names(sample_df)) sample_df$sample <- rownames(sample_df)
sample_df$sample <- as.character(sample_df$sample)
sample_df[[group_col]] <- as.character(sample_df[[group_col]])
label_levels <- base_payload$label_levels

if (!group_col %in% names(sample_df)) {
  stop("group_col not present in cached sample_df: ", group_col)
}

layers_base <- base_payload$layers_base
result_motifs <- unique(stats::na.omit(as.character(unlist(lapply(
  c("edge", "motifs2", "triangle", "wedge"),
  function(layer_name) {
    tbl <- layers_base[[layer_name]]
    if (is.null(tbl) || !is.data.frame(tbl) || !"motif" %in% names(tbl)) character() else tbl$motif
  }
), use.names = FALSE))))
count_motifs <- unique(stats::na.omit(as.character(unlist(lapply(
  c("edge", "triangle", "wedge"),
  function(layer_name) {
    mat <- counts$raw_count[[layer_name]]
    if (is.null(mat)) character() else rownames(mat)
  }
), use.names = FALSE))))
all_motifs <- unique(c(result_motifs, count_motifs))
if (!length(result_motifs)) {
  stop("No tested edge or motif2 results found in base cache. Rerun the base report with the volume-first CellEdgeR API.")
}
if (!length(all_motifs)) {
  stop("No motif counts found in base cache.")
}
filtered_motifs <- unique(stats::na.omit(as.character(base_payload$edger$filter$filtered_motifs)))

nominal_motif_tbl <- layers_base$motifs2
if (is.null(nominal_motif_tbl) || !is.data.frame(nominal_motif_tbl) || !nrow(nominal_motif_tbl)) {
  stop("No tested motif2 table found in base cache.")
}
nominal_motif_tbl <- nominal_motif_tbl %>%
  filter(motif_type %in% c("wedge", "triangle"), is.finite(PValue)) %>%
  arrange(PValue, FDR, desc(abs(logFC)), motif) %>%
  mutate(nominal_rank = row_number())
top_nominal_motif_tbl <- nominal_motif_tbl %>%
  slice_head(n = n_top_nominal_motifs)
if (!nrow(top_nominal_motif_tbl)) {
  stop("No tested W/T motifs with finite nominal p-values found in base cache.")
}

label_palette <- setNames(
  grDevices::hcl.colors(length(label_levels), palette = "Dark 3"),
  label_levels
)
```

```{r helpers}
motif_layer_from_key <- function(motif_key) {
  prefix <- sub("_.*", "", motif_key)
  switch(prefix,
    N = "node",
    E = "edge",
    T = "triangle",
    W = "wedge",
    TP = "triplet",
    TW = "triangle",
    stop("Cannot infer motif layer from key: ", motif_key)
  )
}

resolve_motif_key <- function(query, motif_keys) {
  exact_idx <- which(tolower(motif_keys) == tolower(query))
  if (length(exact_idx)) {
    resolved <- motif_keys[exact_idx[1]]
    return(list(resolved = resolved, edit_distance = 0L))
  }

  search_keys <- motif_keys
  if (grepl("^TP_", query, ignore.case = TRUE)) {
    # TP_* merged triplets are no longer modeled; resolve to the nearest
    # available wedge or triangle key with the same three labels.
    motif2_keys <- motif_keys[grepl("^[TW]_", motif_keys)]
    if (length(motif2_keys)) search_keys <- motif2_keys
  }

  d <- as.vector(utils::adist(query, search_keys, ignore.case = TRUE))
  best_idx <- which.min(d)
  list(resolved = search_keys[best_idx], edit_distance = as.integer(d[best_idx]))
}

get_norm_values <- function(motif_key, offset_mode = "volume") {
  if (!identical(offset_mode, "volume")) {
    stop("Only volume normalized counts are available in the volume-first CellEdgeR API.")
  }

  vals_df <- get_motif_values(counts, motif_key = motif_key, value = "norm")
  missing_samples <- setdiff(sample_df$sample, rownames(vals_df))
  if (length(missing_samples)) {
    stop("Normalized values missing cached samples: ", paste(missing_samples, collapse = ", "))
  }
  vals <- as.numeric(vals_df[sample_df$sample, motif_key])
  names(vals) <- sample_df$sample
  vals
}

select_samples_for_motif <- function(motif_key, offset_mode = "volume", n_per_group = 6, groups = c("ICI", "untreated")) {
  vals <- get_norm_values(motif_key, offset_mode = offset_mode)
  df <- data.frame(
    sample = names(vals),
    normalized_count = as.numeric(vals),
    offset_mode = offset_mode,
    normalization = unname(offset_mode_labels[[offset_mode]]),
    stringsAsFactors = FALSE
  ) %>%
    left_join(sample_df[, c("sample", group_col)], by = "sample") %>%
    rename(group = all_of(group_col)) %>%
    filter(!is.na(group), group %in% groups)

  df$sample_num <- suppressWarnings(as.numeric(df$sample))
  df <- df %>%
    mutate(group = factor(group, levels = groups)) %>%
    arrange(group, desc(normalized_count), sample_num, sample) %>%
    group_by(group) %>%
    slice_head(n = n_per_group) %>%
    mutate(rank_within_group = row_number()) %>%
    ungroup() %>%
    select(offset_mode, normalization, sample, group, rank_within_group, normalized_count)

  df$group <- as.character(df$group)
  df
}

make_sample_plot <- function(motif_key, sample_name, sample_group, normalized_count, offset_mode) {
  plot_sample_graph(
    graphs,
    sample_name = sample_name,
    max_edge_len = edge_cut,
    motif_key = motif_key,
    motif_node_size = 2.4,
    dim_node_nonmotif = 0.45,
    alpha_node_nonmotif = 0.18,
    alpha_edge_nonmotif = 0.18
  ) +
    scale_color_manual(values = label_palette, limits = label_levels, drop = FALSE) +
    ggtitle(sprintf(
      "Sample %s | %s | %s norm=%.3f",
      sample_name,
      sample_group,
      offset_mode_labels[[offset_mode]],
      normalized_count
    )) +
    theme(
      legend.position = "none",
      plot.title = element_text(size = 10)
    )
}

draw_plot_grid <- function(plot_list, ncol = 3) {
  n <- length(plot_list)
  nrow <- max(1, ceiling(n / ncol))
  grid::grid.newpage()
  grid::pushViewport(grid::viewport(layout = grid::grid.layout(nrow, ncol)))
  for (i in seq_along(plot_list)) {
    row <- ceiling(i / ncol)
    col <- ((i - 1) %% ncol) + 1
    print(plot_list[[i]], vp = grid::viewport(layout.pos.row = row, layout.pos.col = col))
  }
  invisible(NULL)
}

panel_slug <- function(motif_key) {
  gsub("[^A-Za-z0-9]+", "_", motif_key)
}

pair_key_local <- function(a, b) {
  pair <- sort(c(a, b))
  paste0("E_", paste(pair, collapse = "_"))
}

submotif_keys_for_motif <- function(motif_key) {
  labels <- strsplit(sub("^[^_]+_", "", motif_key), "_", fixed = TRUE)[[1]]
  layer <- motif_layer_from_key(motif_key)

  if (layer %in% c("triangle", "wedge", "triplet")) {
    if (length(labels) != 3L) {
      stop("Three-label motif does not decode to three labels: ", motif_key)
    }
    sub_keys <- c(
      if (identical(layer, "triplet")) paste0("T_", paste(labels, collapse = "_")) else character(),
      if (identical(layer, "triplet")) paste0("W_", paste(labels, collapse = "_")) else character(),
      pair_key_local(labels[1], labels[2]),
      pair_key_local(labels[1], labels[3]),
      pair_key_local(labels[2], labels[3])
    )
  } else if (identical(layer, "edge")) {
    if (length(labels) != 2L) {
      stop("Edge motif does not decode to two labels: ", motif_key)
    }
    sub_keys <- paste0("N_", labels)
  } else {
    sub_keys <- character()
  }

  unique(sub_keys[sub_keys %in% all_motifs])
}

get_motif_with_submotifs <- function(motif_key, offset_mode = "volume") {
  components <- c(motif_key, setdiff(submotif_keys_for_motif(motif_key), motif_key))
  out <- data.frame(
    sample = sample_df$sample,
    stringsAsFactors = FALSE
  )
  for (component in components) {
    out[[component]] <- as.numeric(get_norm_values(component, offset_mode = offset_mode)[sample_df$sample])
  }

  rownames(out) <- out$sample
  out[, setdiff(colnames(out), "sample"), drop = FALSE]
}

make_submotif_boxplot_df <- function(motif_key, offset_mode = "volume") {
  vals <- get_motif_with_submotifs(motif_key, offset_mode = offset_mode)
  components <- colnames(vals)
  long_df <- data.frame(
    sample = rep(rownames(vals), times = length(components)),
    component = rep(components, each = nrow(vals)),
    normalized_count = as.numeric(as.matrix(vals)),
    offset_mode = offset_mode,
    normalization = unname(offset_mode_labels[[offset_mode]]),
    stringsAsFactors = FALSE
  ) %>%
    left_join(sample_df[, c("sample", group_col)], by = "sample") %>%
    rename(group = all_of(group_col)) %>%
    filter(!is.na(group), group %in% group_order) %>%
    mutate(log2_normalized_count = log2(normalized_count + 1))

  long_df$group <- factor(long_df$group, levels = group_order)
  long_df$component <- factor(long_df$component, levels = components)
  long_df
}

make_submotif_boxplot_panel <- function(motif_key, offset_mode = "volume") {
  plot_df <- make_submotif_boxplot_df(motif_key, offset_mode = offset_mode)

  ggplot(plot_df, aes(x = group, y = log2_normalized_count, fill = group)) +
    geom_boxplot(alpha = 0.7, width = 0.55, outlier.alpha = 0.25) +
    geom_jitter(width = 0.12, alpha = 0.65, size = 1.1, color = "#1b4f72") +
    facet_wrap(~ component, scales = "free_y", ncol = 2) +
    scale_fill_manual(values = c(ICI = "#c65d2e", untreated = "#3b7a57"), drop = FALSE) +
    labs(
      title = paste("Logged normalized counts for motif and submotifs:", motif_key),
      subtitle = offset_mode_labels[[offset_mode]],
      x = NULL,
      y = "log2(normalized count + 1)"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none",
      strip.text = element_text(size = 9),
      plot.title = element_text(size = 11)
    )
}

save_panel_png <- function(motif_key, selected_df, out_path, offset_mode) {
  plot_list <- lapply(seq_len(nrow(selected_df)), function(i) {
    make_sample_plot(
      motif_key = motif_key,
      sample_name = selected_df$sample[i],
      sample_group = selected_df$group[i],
      normalized_count = selected_df$normalized_count[i],
      offset_mode = offset_mode
    )
  })

  grDevices::png(
    filename = out_path,
    width = 1800,
    height = 2400,
    res = 180
  )
  on.exit(grDevices::dev.off(), add = TRUE)
  draw_plot_grid(plot_list, ncol = 3)
  invisible(out_path)
}

save_boxplot_png <- function(motif_key, out_path, offset_mode) {
  p <- make_submotif_boxplot_panel(motif_key, offset_mode = offset_mode)
  grDevices::png(
    filename = out_path,
    width = 1800,
    height = 1200,
    res = 180
  )
  on.exit(grDevices::dev.off(), add = TRUE)
  print(p)
  invisible(out_path)
}

resolved_tbl <- top_nominal_motif_tbl %>%
  transmute(
    selection_rank = nominal_rank,
    resolved_motif = motif,
    motif_layer = motif_type,
    logFC = logFC,
    PValue = PValue,
    FDR = FDR,
    edgeR_test_status = "tested"
  )

selected_samples <- setNames(
  lapply(offset_modes, function(offset_mode) {
    setNames(
      lapply(
        resolved_tbl$resolved_motif,
        select_samples_for_motif,
        offset_mode = offset_mode,
        n_per_group = n_per_group,
        groups = group_order
      ),
      resolved_tbl$resolved_motif
    )
  }),
  offset_modes
)

offset_path_slug <- function(offset_mode) {
  offset_mode
}

panel_rel_path_for_mode <- function(motif_key, offset_mode) {
  file.path("hochschulz_2022_melanoma_selected_motifs_panels", paste0(panel_slug(motif_key), ".png"))
}

panel_paths <- setNames(lapply(offset_modes, function(mode) setNames(vector("list", nrow(resolved_tbl)), resolved_tbl$resolved_motif)), offset_modes)
boxplot_paths <- setNames(lapply(offset_modes, function(mode) setNames(vector("list", nrow(resolved_tbl)), resolved_tbl$resolved_motif)), offset_modes)

for (offset_mode in offset_modes) {
  for (motif_key in resolved_tbl$resolved_motif) {
    panel_rel_path <- panel_rel_path_for_mode(motif_key, offset_mode)
    panel_abs_path <- file.path(report_dir, panel_rel_path)
    dir.create(dirname(panel_abs_path), recursive = TRUE, showWarnings = FALSE)
    if (force_render_panels || !file.exists(panel_abs_path)) {
      save_panel_png(motif_key, selected_samples[[offset_mode]][[motif_key]], panel_abs_path, offset_mode = offset_mode)
    }
    panel_paths[[offset_mode]][[motif_key]] <- panel_rel_path

    boxplot_rel_path <- file.path(
      "hochschulz_2022_melanoma_selected_motifs_boxplots_logged",
      offset_path_slug(offset_mode),
      paste0(panel_slug(motif_key), ".png")
    )
    boxplot_abs_path <- file.path(report_dir, boxplot_rel_path)
    dir.create(dirname(boxplot_abs_path), recursive = TRUE, showWarnings = FALSE)
    if (force_render_panels || !file.exists(boxplot_abs_path)) {
      save_boxplot_png(motif_key, boxplot_abs_path, offset_mode = offset_mode)
    }
    boxplot_paths[[offset_mode]][[motif_key]] <- boxplot_rel_path
  }
}

group_counts <- table(sample_df[[group_col]])
selection_rows <- bind_rows(lapply(offset_modes, function(offset_mode) {
  bind_rows(lapply(names(selected_samples[[offset_mode]]), function(motif_key) {
    dplyr::mutate(selected_samples[[offset_mode]][[motif_key]], motif = motif_key, .before = 1)
  }))
}))
```

## Summary

- Contrast cache: `r basename(base_cache)` using `r group_col`.
- Available samples in cache: `r paste(sprintf("%s = %d", names(group_counts), as.integer(group_counts)), collapse = ", ")`.
- Normalization view: `r paste(unname(offset_mode_labels[offset_modes]), collapse = " and ")`.
- Selected motifs: top `r nrow(resolved_tbl)` tested wedge/triangle motifs by nominal `PValue`.
- Per motif, selection rule: top `r n_per_group` `ICI` samples and top `r n_per_group` `untreated` samples by volume-normalized motif abundance.
- Boxplots show `log2(normalized count + 1)`.
- Low-count motifs filtered by `edgeR::filterByExpr()` are not eligible for this nominal-p-value selection.

```{r motif-map}
knitr::kable(resolved_tbl, align = "l")
```

```{r selection-preview}
knitr::kable(selection_rows, digits = 4, align = "l")
```

```{r motif-sections, results='asis'}
for (i in seq_len(nrow(resolved_tbl))) {
  motif_info <- resolved_tbl[i, , drop = FALSE]
  motif_key <- motif_info$resolved_motif[[1]]

  cat("## ", motif_key, "\n\n", sep = "")
  cat("- Nominal PValue rank: `", motif_info$selection_rank[[1]], "`\n", sep = "")
  cat("- Layer: `", motif_info$motif_layer[[1]], "`\n", sep = "")
  cat("- logFC: `", signif(motif_info$logFC[[1]], 4), "`\n", sep = "")
  cat("- Nominal PValue: `", signif(motif_info$PValue[[1]], 4), "`\n", sep = "")
  cat("- FDR: `", signif(motif_info$FDR[[1]], 4), "`\n", sep = "")
  cat("- edgeR test status: `", motif_info$edgeR_test_status[[1]], "`\n", sep = "")
  cat("- Selection: highest volume-normalized motif abundance within each group.\n\n")

  for (offset_mode in offset_modes) {
    samples_tbl <- selected_samples[[offset_mode]][[motif_key]]
    panel_rel_path <- panel_paths[[offset_mode]][[motif_key]]
    boxplot_rel_path <- boxplot_paths[[offset_mode]][[motif_key]]

    cat("### ", offset_mode_labels[[offset_mode]], "\n\n", sep = "")
    print(knitr::kable(samples_tbl, digits = 4, align = "l"))
    cat("\n\n")
    cat("Spatial motif plots selected by ", offset_mode_labels[[offset_mode]], " normalized abundance:\n\n", sep = "")
    cat("![](", panel_rel_path, ")\n\n", sep = "")
    cat("Logged normalized motif and submotif boxplots:\n\n")
    cat("![](", boxplot_rel_path, ")\n\n", sep = "")
  }
}
```
