---
title: "HochSchulz_2022_Melanoma base cache"
output:
  html_document:
    code_folding: hide
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE)

repo_candidates <- unique(normalizePath(c(
  getwd(),
  file.path(getwd(), ".."),
  file.path(getwd(), "..", ".."),
  file.path(getwd(), "..", "..", "..")
), mustWork = FALSE))
repo_hit <- vapply(repo_candidates, function(path) {
  desc <- file.path(path, "DESCRIPTION")
  file.exists(desc) && any(grepl("^Package: CellEdgeR$", readLines(desc)))
}, logical(1))
root_dir <- if (any(repo_hit)) repo_candidates[which(repo_hit)[1]] else NA_character_
use_local <- !is.na(root_dir)
```

```{r libs}
if (use_local && requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(root_dir)
} else if (!requireNamespace("CellEdgeR", quietly = TRUE)) {
  stop("Install CellEdgeR or run this report inside the repo with devtools::load_all().")
}
if (!requireNamespace("imcdatasets", quietly = TRUE)) {
  stop("Install imcdatasets with BiocManager::install('imcdatasets').")
}
if (!requireNamespace("SpatialExperiment", quietly = TRUE)) {
  stop("Install SpatialExperiment.")
}
if (!requireNamespace("SummarizedExperiment", quietly = TRUE)) {
  stop("Install SummarizedExperiment.")
}

library(CellEdgeR)
library(imcdatasets)
library(SpatialExperiment)
library(SummarizedExperiment)
library(dplyr)
library(tidyr)
library(ggplot2)

helpers_candidates <- c(
  if (use_local) file.path(root_dir, "analysis", "hochchulz_2022_melanoma", "hochschulz_2022_melanoma_shared.R"),
  "analysis/hochchulz_2022_melanoma/hochschulz_2022_melanoma_shared.R",
  "hochschulz_2022_melanoma_shared.R"
)
helpers_candidates <- unique(helpers_candidates[!is.na(helpers_candidates)])
helpers_hits <- helpers_candidates[file.exists(helpers_candidates)]
if (!length(helpers_hits)) {
  stop("Missing helper script. Tried: ", paste(helpers_candidates, collapse = ", "))
}
helpers_path <- helpers_hits[1]
source(helpers_path)

if (!exists("top_motifs2", where = asNamespace("CellEdgeR"), inherits = FALSE)) {
  stop("This report requires a volume-first CellEdgeR version with top_motifs2().")
}
```

```{r params}
seed <- 1
set.seed(seed)
full_dataset <- FALSE
spe_type <- "spe"
label_col <- "cell_type"
image_col <- "image_name"
x_col <- "cell_x"
y_col <- "cell_y"
group_col <- "patient_treatment_group_before_surgery"
group_levels <- c("untreated", "ICI")
min_cells_per_image <- 200
max_cells_per_image <- NULL
max_images <- NULL
include_wedge <- TRUE
edge_len_quantile <- 0.99
n_cores <- 6

cache_dir <- if (use_local) file.path(root_dir, "analysis", "cache") else tempdir()
if (!dir.exists(cache_dir)) dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
base_cache <- file.path(cache_dir, "hochschulz_base.rds")
```

```{r data}
spe <- load_hochschulz(data_type = spe_type, full_dataset = full_dataset)
cd <- as.data.frame(colData(spe))

if (!group_col %in% names(cd)) {
  stop("group_col not found in colData(spe): ", group_col)
}
if (!all(group_levels %in% unique(as.character(cd[[group_col]])))) {
  stop("Requested group_levels are not all present in colData(spe): ",
       paste(group_levels, collapse = ", "))
}
```

```{r build}
prep <- prepare_cells_by_sample(
  spe,
  label_col = label_col,
  image_col = image_col,
  x_col = x_col,
  y_col = y_col,
  min_cells = min_cells_per_image,
  max_cells = max_cells_per_image,
  max_images = max_images
)

cells_by_sample <- prep$cells_by_sample
cell_df <- prep$cell_df
cell_df$image <- as.character(cell_df$image)
names(cells_by_sample) <- as.character(names(cells_by_sample))

sample_df <- data.frame(sample = names(cells_by_sample), stringsAsFactors = FALSE)
group_map <- as.data.frame(colData(spe)) %>%
  select(all_of(c(image_col, group_col))) %>%
  distinct() %>%
  group_by(.data[[image_col]]) %>%
  slice(1) %>%
  ungroup()
names(group_map)[names(group_map) == image_col] <- "sample"
group_map$sample <- as.character(group_map$sample)
sample_df$sample <- as.character(sample_df$sample)
sample_df <- sample_df %>% left_join(group_map, by = "sample")
rownames(sample_df) <- sample_df$sample
full_group_counts <- table(sample_df[[group_col]], useNA = "ifany")
keep_samples <- !is.na(sample_df[[group_col]]) & sample_df[[group_col]] %in% group_levels
sample_df <- sample_df[keep_samples, , drop = FALSE]
cells_by_sample <- cells_by_sample[rownames(sample_df)]
cell_df <- cell_df[cell_df$image %in% rownames(sample_df), , drop = FALSE]
sample_df[[group_col]] <- factor(as.character(sample_df[[group_col]]), levels = group_levels)

group_counts <- table(sample_df[[group_col]])
excluded_group_counts <- full_group_counts[!names(full_group_counts) %in% group_levels]
label_levels <- sort(unique(as.character(cell_df$label)))
label_levels_md <- paste(sprintf("`%s`", label_levels), collapse = ", ")
group_counts_md <- paste(sprintf("`%s` = %d", names(group_counts), as.integer(group_counts)), collapse = ", ")
excluded_counts_md <- if (length(excluded_group_counts)) {
  paste(sprintf("`%s` = %d", names(excluded_group_counts), as.integer(excluded_group_counts)), collapse = ", ")
} else {
  "`none`"
}
design_preview <- stats::model.matrix(as.formula(paste0("~ ", group_col)), data = sample_df)
coef_candidates <- setdiff(colnames(design_preview), "(Intercept)")
if (length(coef_candidates) != 1) {
  stop("Expected exactly one non-intercept coefficient after filtering to two group levels.")
}
coef_name <- coef_candidates[1]
contrast_label <- paste0(levels(sample_df[[group_col]])[2], " vs ", levels(sample_df[[group_col]])[1])
```

## Analysis Summary

- Purpose: exploratory validation contrast, not a causal or predictive claim.
- Dataset: HochSchulz 2022 Melanoma from `imcdatasets` (`full_dataset = FALSE`, one graph per `r image_col`).
- Differential motif contrast: `r group_col`, coefficient `r coef_name` = `r contrast_label`.
- logFC direction: positive = higher in `r levels(sample_df[[group_col]])[2]`; negative = higher in `r levels(sample_df[[group_col]])[1]`.
- Images analyzed: `r nrow(sample_df)` (`r group_counts_md`).
- Images excluded from this contrast: `r excluded_counts_md`.
- Cell label column: `r label_col`.
- Labels used (`r length(label_levels)`): `r label_levels_md`.
- Total cells analyzed: `r nrow(cell_df)`.

```{r graphs}
graphs <- build_cell_graphs(cells_by_sample, n_cores = n_cores, verbose = TRUE)
edge_len <- unlist(lapply(graphs$per_sample_graph, function(ps) ps$edge_len))
edge_cut <- as.numeric(stats::quantile(edge_len, probs = edge_len_quantile, na.rm = TRUE))

counts <- count_motifs_graphs(
  graph_obj = graphs,
  max_edge_len = edge_cut,
  include_wedge = include_wedge,
  n_cores = n_cores,
  verbose = TRUE
)
```

```{r models}
res <- motif_edger(
  cellgraph = counts,
  sample_df = sample_df,
  design_formula = as.formula(paste0("~ ", group_col)),
  verbose = TRUE
)
layers_base <- collect_layer_tables(res, coef_name = coef_name)
```

```{r volcano}
edges_tbl <- layers_base$edge
motifs2_tbl <- layers_base$motifs2
```

## Volcano Plots

- Points are colored by FDR <= 0.05.
- Positive logFC: enriched in `r levels(sample_df[[group_col]])[2]`.
- Negative logFC: enriched in `r levels(sample_df[[group_col]])[1]`.

```{r volcano_plots}
plot_motif_volcano(
  edges_tbl,
  title = "Edges: volume offsets",
  label_mode = "top_bottom_n",
  label_n = 6
)

plot_motif_volcano(
  motifs2_tbl,
  title = "Motif2: wedges and triangles, volume offsets",
  label_mode = "top_bottom_n",
  label_n = 6
)

plot_motif_volcano_triptych(
  motifs2_tbl,
  n_side = 5,
  title = "Motif2: depleted vs enriched",
  subtitle = paste0(group_col, " (", contrast_label, ")")
)
```

```{r cache}
base_payload <- list(
  params = list(
    seed = seed,
    full_dataset = full_dataset,
    spe_type = spe_type,
    label_col = label_col,
    image_col = image_col,
    x_col = x_col,
    y_col = y_col,
    group_col = group_col,
    group_levels = group_levels,
    contrast_label = contrast_label,
    min_cells_per_image = min_cells_per_image,
    max_cells_per_image = max_cells_per_image,
    max_images = max_images,
    include_wedge = include_wedge,
    edge_len_quantile = edge_len_quantile,
    n_cores = n_cores
  ),
  graphs = graphs,
  counts = counts,
  edge_cut = edge_cut,
  sample_df = sample_df,
  cells_by_sample = cells_by_sample,
  cell_df = cell_df,
  label_levels = label_levels,
  coef_name = coef_name,
  edger = list(
    result_sets = res$edger$result_sets,
    statistics = res$edger$statistics,
    offset_version = res$edger$offset_version,
    filter = res$edger$filter
  ),
  layers_base = layers_base
)

saveRDS(base_payload, base_cache)
message("Saved base cache: ", base_cache)
```

Summary: `r nrow(sample_df)` samples, `r nrow(cell_df)` cells, `r length(unique(cell_df$label))` labels, edge length cutoff = `r signif(edge_cut, 3)`.
