---
title: "CellEdgeR validation on HochSchulz_2022_Melanoma"
output:
  html_document:
    code_folding: hide
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE)

if(!require("imcdatasets")){
  BiocManager::install("terra")
  BiocManager::install("cytomapper")
  BiocManager::install('imcdatasets')
}

repo_candidates <- unique(normalizePath(c(
  getwd(),
  file.path(getwd(), ".."),
  file.path(getwd(), "..", ".."),
  file.path(getwd(), "..", "..", "..")
), mustWork = FALSE))
repo_hit <- vapply(repo_candidates, function(path) {
  desc <- file.path(path, "DESCRIPTION")
  file.exists(desc) && any(grepl("^Package: CellEdgeR$", readLines(desc)))
}, logical(1))
root_dir <- if (any(repo_hit)) repo_candidates[which(repo_hit)[1]] else NA_character_
use_local <- !is.na(root_dir)

```

# Overview

This report validates CellEdgeR on the HochSchulz_2022_Melanoma dataset from the
`imcdatasets` package, using volume-based Chung-Lu-style offsets for edges,
wedges, and triangles. It also includes negative-control label shuffles
(QQ/histograms of p-values) and stability checks (sample downsampling and
spatial cropping).

# Parameters

```{r params}
# Reproducibility

set.seed(1)

# Reproducibility
seed <- 1
set.seed(seed)

# Dataset retrieval
full_dataset <- FALSE  # FALSE loads the 50-image subset used by imcdatasets
spe_type <- "spe"       # use SpatialExperiment

# Columns to use (override if needed)
label_col <- "cell_type"
image_col <- "image_name"
x_col <- "cell_x"
y_col <- "cell_y"

# Sample-level grouping (override after inspection)
# group_col <- "patient_treatment_group_after_surgery"
# group_col <- "patient_status_at_3months"
# group_col <- "sample_tissue_type"
group_col <- NULL

# Filtering / runtime controls
min_cells_per_image <- 200
max_cells_per_image <- NULL  # set to integer to downsample per image
max_images <- NULL           # set to integer to sample images

# Graph / counting
include_wedge <- TRUE
edge_len_quantile <- 0.99
n_cores <- 6

# Parallelism note: mclapply is used when n_cores > 1 on Unix-alikes.

# Negative-control label shuffle (used by permutation script)
n_perm <- 20
perm_seed <- 100

# Stability checks (used by stability script)
sample_keep_fracs <- c(0.8, 0.6)
sample_keep_reps <- 5
spatial_keep_fracs <- c(0.9, 0.7, 0.5)
top_n_stability <- 20

# Cache (used by separate permutation/stability scripts)
write_cache <- TRUE
cache_dir <- if (use_local) file.path(root_dir, "analysis", "cache") else tempdir()
```

# Libraries

```{r libs}
loaded_dev <- FALSE
if (use_local && requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(root_dir)
  loaded_dev <- TRUE
} else if (!requireNamespace("CellEdgeR", quietly = TRUE)) {
  stop("Install CellEdgeR or run this report inside the repo with devtools::load_all().")
}
if (!requireNamespace("imcdatasets", quietly = TRUE)) {
  stop("Install imcdatasets with BiocManager::install('imcdatasets').")
}
if (!requireNamespace("SpatialExperiment", quietly = TRUE)) {
  stop("Install SpatialExperiment.")
}

if (!loaded_dev) library(CellEdgeR)
library(imcdatasets)
library(SpatialExperiment)
library(SummarizedExperiment)
library(dplyr)
library(tidyr)
library(ggplot2)

if (!exists("top_motifs2", where = asNamespace("CellEdgeR"), inherits = FALSE)) {
  stop("This report requires a volume-first CellEdgeR version with top_motifs2(). ",
       "Install from the repo or run devtools::load_all().")
}
```

# Helper functions

```{r helpers}
helpers_path <- if (use_local) {
  file.path(root_dir, "analysis", "hochchulz_2022_melanoma", "hochschulz_2022_melanoma_shared.R")
} else {
  "analysis/hochchulz_2022_melanoma/hochschulz_2022_melanoma_shared.R"
}
if (!file.exists(helpers_path)) {
  stop("Missing helper script at: ", helpers_path)
}
source(helpers_path)
```

# Load dataset

```{r data}
spe <- load_hochschulz(data_type = spe_type, full_dataset = full_dataset)

cd <- as.data.frame(colData(spe))

# Auto-detect group column if not specified
if (is.null(group_col)) {
  group_candidates <- c(
    "patient_treatment_group_after_surgery",
    "patient_treatment_group_before_surgery",
    "patient_status_at_3months",
    "patient_relapse",
    "sample_tissue_type",
    "sample_location",
    "patient_mutation"
  )
  group_col <- pick_group_col(cd, group_candidates)
}

if (is.na(group_col)) {
  stop("Set group_col explicitly after inspecting colData(spe).")
}
```

# Build sample list

```{r build}
prep <- prepare_cells_by_sample(
  spe,
  label_col = label_col,
  image_col = image_col,
  x_col = x_col,
  y_col = y_col,
  min_cells = min_cells_per_image,
  max_cells = max_cells_per_image,
  max_images = max_images
)

cells_by_sample <- prep$cells_by_sample
cell_df <- prep$cell_df
cell_df$image <- as.character(cell_df$image)
names(cells_by_sample) <- as.character(names(cells_by_sample))

# sample-level metadata aligned to cells_by_sample
sample_df <- data.frame(sample = names(cells_by_sample), stringsAsFactors = FALSE)
group_map <- as.data.frame(colData(spe)) %>%
  select(all_of(c(image_col, group_col))) %>%
  distinct() %>%
  group_by(.data[[image_col]]) %>%
  slice(1) %>%
  ungroup()
names(group_map)[names(group_map) == image_col] <- "sample"
group_map$sample <- as.character(group_map$sample)
sample_df$sample <- as.character(sample_df$sample)
sample_df <- sample_df %>% left_join(group_map, by = "sample")

rownames(sample_df) <- sample_df$sample

# ensure group is a factor
sample_df[[group_col]] <- factor(sample_df[[group_col]])

# drop samples with missing group assignment
missing_group <- is.na(sample_df[[group_col]])
if (any(missing_group)) {
  drop_samples <- rownames(sample_df)[missing_group]
  message("Dropping samples with missing group: ", paste(drop_samples, collapse = ", "))
  sample_df <- sample_df[!missing_group, , drop = FALSE]
  cells_by_sample <- cells_by_sample[rownames(sample_df)]
  cell_df <- cell_df[cell_df$image %in% rownames(sample_df), , drop = FALSE]
}

sample_df
```

# Basic summaries

```{r summaries}
# cells per image
cell_counts <- cell_df %>% count(image, name = "n_cells")

# label distribution
label_counts <- cell_df %>% count(label, name = "n_cells") %>% arrange(desc(n_cells))

p_cells <- ggplot(cell_counts, aes(x = n_cells)) +
  geom_histogram(bins = 30, fill = "#4477aa", color = "white") +
  labs(title = "Cells per image", x = "Cells", y = "Images") +
  theme_minimal()

p_labels <- ggplot(label_counts, aes(x = reorder(label, n_cells), y = n_cells)) +
  geom_col(fill = "#aa3377") +
  coord_flip() +
  labs(title = "Cell-type counts", x = NULL, y = "Cells") +
  theme_minimal()

p_cells
p_labels
```

# Build graphs and count motifs

```{r count}
graphs <- build_cell_graphs(cells_by_sample, n_cores = n_cores, verbose = TRUE)

# Edge length distribution to pick a pruning threshold
edge_len <- unlist(lapply(graphs$per_sample_graph, function(ps) ps$edge_len))
edge_cut <- as.numeric(stats::quantile(edge_len, probs = edge_len_quantile, na.rm = TRUE))

p_edge_len <- ggplot(data.frame(edge_len = edge_len), aes(edge_len)) +
  geom_density(fill = "#44aa77", alpha = 0.4) +
  scale_x_log10() +
  labs(title = "Edge length distribution (log10)", x = "Edge length", y = "Density") +
  theme_minimal()

p_edge_len

counts <- count_motifs_graphs(
  graph_obj = graphs,
  max_edge_len = edge_cut,
  include_wedge = include_wedge,
  n_cores = n_cores,
  verbose = TRUE
)
```

# Fit models

```{r models}
# Use the first non-intercept coefficient by default
coef_name <- NULL

res <- motif_edger(
  cellgraph = counts,
  sample_df = sample_df,
  design_formula = as.formula(paste0("~ ", group_col)),
  verbose = TRUE
)

layers_base <- collect_layer_tables(res, coef_name = coef_name)
edges_tbl <- layers_base$edge
motifs2_tbl <- layers_base$motifs2
```

```{r cache-base}
if (isTRUE(write_cache)) {
  if (!dir.exists(cache_dir)) dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
  base_cache <- file.path(cache_dir, "hochschulz_validation_base.rds")
  base_payload <- list(
    params = list(
      seed = seed,
      full_dataset = full_dataset,
      spe_type = spe_type,
      label_col = label_col,
      image_col = image_col,
      x_col = x_col,
      y_col = y_col,
      group_col = group_col,
      min_cells_per_image = min_cells_per_image,
      max_cells_per_image = max_cells_per_image,
      max_images = max_images,
      include_wedge = include_wedge,
      edge_len_quantile = edge_len_quantile,
      n_cores = n_cores
    ),
    graphs = graphs,
    counts = counts,
    edge_cut = edge_cut,
    sample_df = sample_df,
    cells_by_sample = cells_by_sample,
    cell_df = cell_df,
    coef_name = coef_name,
    edger = list(
      result_sets = res$edger$result_sets,
      statistics = res$edger$statistics,
      offset_version = res$edger$offset_version
    ),
    layers_base = layers_base
  )
  saveRDS(base_payload, base_cache)
  message("Saved base cache: ", base_cache)
}
```

# Results plots

```{r results-plots}
plot_volcano(edges_tbl, "Edges: volume offsets")
plot_volcano(motifs2_tbl, "Motif2: wedges and triangles, volume offsets")
```

Significant motifs (FDR < 0.05):
- Edges: `r sum(edges_tbl$FDR < 0.05, na.rm = TRUE)` of `r nrow(edges_tbl)`
- Motif2 (wedges and triangles): `r sum(motifs2_tbl$FDR < 0.05, na.rm = TRUE)` of `r nrow(motifs2_tbl)`

# Top motifs

```{r top-motifs}
head(edges_tbl, 10)
head(motifs2_tbl, 10)
```

# Visualizing significant motifs

```{r motif-select}
pick_top <- function(tbl, n = 2) {
  if (!nrow(tbl)) return(character())
  if (all(is.na(tbl$FDR))) {
    tbl <- tbl[order(tbl$PValue, na.last = TRUE), , drop = FALSE]
  } else {
    tbl <- tbl[order(tbl$FDR, tbl$PValue, na.last = TRUE), , drop = FALSE]
  }
  head(tbl$motif, n)
}

edge_motifs <- pick_top(edges_tbl, n = 2)
motifs2_top <- pick_top(motifs2_tbl, n = 2)

edge_motifs
motifs2_top
```

```{r motif-boxplots}
plot_motif_box <- function(motif_key, title) {
  vals <- get_motif_values(counts, motif_key = motif_key, value = "norm")
  df <- data.frame(sample = rownames(vals), value = vals[[motif_key]], stringsAsFactors = FALSE)
  df <- merge(df, sample_df, by.x = "sample", by.y = "row.names", all.x = TRUE)
  ggplot(df, aes_string(x = group_col, y = "value")) +
    geom_boxplot(outlier.alpha = 0.3) +
    geom_jitter(width = 0.15, alpha = 0.5, size = 1) +
    labs(title = title, x = group_col, y = "Normalized count (volume offset)") +
    theme_minimal()
}

for (m in edge_motifs) print(plot_motif_box(m, paste("Edge motif", m)))
for (m in motifs2_top) print(plot_motif_box(m, paste("Motif2", m)))
```

```{r motif-spatial}
samples_show <- sample_df %>%
  group_by(.data[[group_col]]) %>%
  slice_head(n = 1) %>%
  pull(sample)

plot_motif_samples <- function(motif_key, title_prefix) {
  for (s in samples_show) {
    p <- plot_sample_graph(
      graphs,
      sample_name = s,
      max_edge_len = edge_cut,
      motif_key = motif_key,
      motif_node_size = 3,
      dim_node_nonmotif = 0.5,
      alpha_node_nonmotif = 0.2,
      alpha_edge_nonmotif = 0.2
    )
    print(p + ggtitle(paste(title_prefix, motif_key, "sample:", s)))
  }
}

for (m in edge_motifs) plot_motif_samples(m, "Edge")
for (m in motifs2_top) plot_motif_samples(m, "Motif2")
```

# External analyses (run separately)

Long-running analyses are split into standalone scripts that read the base cache and
save their own results. The base cache is written by this Rmd (or via the helper
script below).

- `analysis/hochschulz_2022_melanoma_base.Rmd` (builds `analysis/cache/hochschulz_base.rds`)
- `analysis/hochschulz_2022_melanoma_permutation.Rmd`
- `analysis/hochschulz_2022_melanoma_stability.Rmd`

If the result caches exist, the plots below will be rendered. Otherwise, run the
scripts first.

```{r external-results}
perm_cache <- file.path(cache_dir, "hochschulz_perm_volume_filter_v3.rds")
if (file.exists(perm_cache)) {
  perm <- readRDS(perm_cache)
  perm_pvals <- perm$perm_pvals
  if (nrow(perm_pvals)) {
    qq_df <- bind_rows(lapply(split(perm_pvals, perm_pvals$layer), function(df) {
      make_qq_df(df$PValue, unique(df$layer))
    }))
    p_qq <- ggplot(qq_df, aes(x = exp, y = obs)) +
      geom_point(alpha = 0.4, size = 0.6) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
      facet_wrap(~ layer, scales = "free") +
      labs(title = "Label-shuffle QQ plots (null p-value calibration)",
           x = "Expected -log10(p)", y = "Observed -log10(p)") +
      theme_minimal()
    p_qq

    p_hist <- ggplot(perm_pvals, aes(x = PValue)) +
      geom_histogram(bins = 40, fill = "#4477aa", color = "white") +
      facet_wrap(~ layer, scales = "free_y") +
      labs(title = "Label-shuffle p-value histograms",
           x = "P-value", y = "Count") +
      theme_minimal()
    p_hist

    perm_summary <- perm_pvals %>%
      group_by(layer) %>%
      summarise(
        n = n(),
        prop_p05 = mean(PValue < 0.05, na.rm = TRUE),
        .groups = "drop"
      )
    perm_summary
  }
} else {
  message("Permutation cache not found: ", perm_cache)
}

stability_cache <- file.path(cache_dir, "hochschulz_stability_volume_filter_v3.rds")
if (file.exists(stability_cache)) {
  stab <- readRDS(stability_cache)
  sample_summary <- stab$sample_summary
  sample_pairs <- stab$sample_pairs
  space_summary <- stab$space_summary
  space_pairs <- stab$space_pairs
  random_summary <- stab$random_summary
  random_pairs <- stab$random_pairs

  if (!is.null(sample_summary) && nrow(sample_summary)) {
    cor_long <- sample_summary %>%
      pivot_longer(cols = c(cor_logFC, cor_logP), names_to = "metric", values_to = "value") %>%
      mutate(metric = recode(metric, cor_logFC = "logFC", cor_logP = "-log10P"))

    p_sample_cor <- ggplot(cor_long, aes(x = factor(frac), y = value)) +
      geom_boxplot(outlier.alpha = 0.3) +
      facet_grid(metric ~ layer) +
      labs(title = "Sample downsampling: correlation to baseline",
           x = "Fraction of samples kept (stratified by group)", y = "Correlation") +
      theme_minimal()

    p_sample_top <- ggplot(sample_summary, aes(x = factor(frac), y = top_jaccard)) +
      geom_boxplot(outlier.alpha = 0.3) +
      facet_wrap(~ layer) +
      labs(title = paste0("Sample downsampling: top-", top_n_stability, " motif overlap"),
           x = "Fraction of samples kept", y = "Jaccard overlap") +
      theme_minimal()

    p_sample_cor
    p_sample_top
  }

  if (!is.null(sample_pairs) && nrow(sample_pairs)) {
    min_frac <- min(sample_pairs$frac, na.rm = TRUE)
    pair_plot <- sample_pairs %>% filter(frac == min_frac)
    pair_long <- bind_rows(
      pair_plot %>% transmute(layer, frac, metric = "logFC", base = logFC_base, alt = logFC_alt),
      pair_plot %>% transmute(layer, frac, metric = "-log10P", base = logP_base, alt = logP_alt)
    )
    p_sample_scatter <- ggplot(pair_long, aes(x = base, y = alt)) +
      geom_point(alpha = 0.3, size = 0.6) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
      facet_grid(metric ~ layer, scales = "free") +
      labs(title = paste0("Sample downsampling scatter (keep ", min_frac, ")"),
           x = "Baseline", y = "Downsampled") +
      theme_minimal()
    p_sample_scatter
  }

  if (!is.null(space_summary) && nrow(space_summary)) {
    cor_long <- space_summary %>%
      pivot_longer(cols = c(cor_logFC, cor_logP), names_to = "metric", values_to = "value") %>%
      mutate(metric = recode(metric, cor_logFC = "logFC", cor_logP = "-log10P"))

    p_space_cor <- ggplot(cor_long, aes(x = keep_frac, y = value)) +
      geom_point() +
      geom_line() +
      facet_grid(metric ~ layer) +
      labs(title = "Spatial cropping: correlation to baseline",
           x = "Fraction of spatial area kept", y = "Correlation") +
      theme_minimal()

    p_space_top <- ggplot(space_summary, aes(x = keep_frac, y = top_jaccard)) +
      geom_point() +
      geom_line() +
      facet_wrap(~ layer) +
      labs(title = paste0("Spatial cropping: top-", top_n_stability, " motif overlap"),
           x = "Fraction of spatial area kept", y = "Jaccard overlap") +
      theme_minimal()

    p_space_cor
    p_space_top
  }

  if (!is.null(space_pairs) && nrow(space_pairs)) {
    min_frac <- min(space_pairs$keep_frac, na.rm = TRUE)
    pair_plot <- space_pairs %>% filter(keep_frac == min_frac)
    pair_long <- bind_rows(
      pair_plot %>% transmute(layer, keep_frac, metric = "logFC", base = logFC_base, alt = logFC_alt),
      pair_plot %>% transmute(layer, keep_frac, metric = "-log10P", base = logP_base, alt = logP_alt)
    )
    p_space_scatter <- ggplot(pair_long, aes(x = base, y = alt)) +
      geom_point(alpha = 0.3, size = 0.6) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
      facet_grid(metric ~ layer, scales = "free") +
      labs(title = paste0("Spatial cropping scatter (keep ", min_frac, ")"),
           x = "Baseline", y = "Cropped") +
      theme_minimal()
    p_space_scatter
  }

  if (!is.null(random_summary) && nrow(random_summary)) {
    cor_long <- random_summary %>%
      pivot_longer(cols = c(cor_logFC, cor_logP), names_to = "metric", values_to = "value") %>%
      mutate(metric = recode(metric, cor_logFC = "logFC", cor_logP = "-log10P"))

    p_rand_cor <- ggplot(cor_long, aes(x = keep_frac, y = value)) +
      geom_point() +
      geom_line() +
      facet_grid(metric ~ layer) +
      labs(title = "Random cell downsampling: correlation to baseline",
           x = "Fraction of cells kept", y = "Correlation") +
      theme_minimal()

    p_rand_top <- ggplot(random_summary, aes(x = keep_frac, y = top_jaccard)) +
      geom_point() +
      geom_line() +
      facet_wrap(~ layer) +
      labs(title = paste0("Random cell downsampling: top-", top_n_stability, " motif overlap"),
           x = "Fraction of cells kept", y = "Jaccard overlap") +
      theme_minimal()

    p_rand_cor
    p_rand_top
  }

  if (!is.null(random_pairs) && nrow(random_pairs)) {
    min_frac <- min(random_pairs$keep_frac, na.rm = TRUE)
    pair_plot <- random_pairs %>% filter(keep_frac == min_frac)
    pair_long <- bind_rows(
      pair_plot %>% transmute(layer, keep_frac, metric = "logFC", base = logFC_base, alt = logFC_alt),
      pair_plot %>% transmute(layer, keep_frac, metric = "-log10P", base = logP_base, alt = logP_alt)
    )
    p_rand_scatter <- ggplot(pair_long, aes(x = base, y = alt)) +
      geom_point(alpha = 0.3, size = 0.6) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
      facet_grid(metric ~ layer, scales = "free") +
      labs(title = paste0("Random cell downsampling scatter (keep ", min_frac, ")"),
           x = "Baseline", y = "Downsampled") +
      theme_minimal()
    p_rand_scatter
  }
} else {
  message("Stability cache not found: ", stability_cache)
}
```

# External analyses summary

```{r external-summary, results='asis'}
perm_pvals <- get0("perm_pvals", ifnotfound = data.frame())
sample_summary <- get0("sample_summary", ifnotfound = data.frame())
space_summary <- get0("space_summary", ifnotfound = data.frame())
random_summary <- get0("random_summary", ifnotfound = data.frame())

format_layer_min <- function(df) {
  if (!nrow(df)) return("none")
  mins <- tapply(df$n, df$layer, function(x) min(x, na.rm = TRUE))
  paste(names(mins), unname(mins), sep = ": ", collapse = "; ")
}

cat("Permutation p-values: total rows = ",
    if (nrow(perm_pvals)) nrow(perm_pvals) else 0,
    "; layers = ",
    if (nrow(perm_pvals)) length(unique(perm_pvals$layer)) else 0,
    ".\n\n", sep = "")
cat("Sample downsampling min motif counts per layer: ",
    format_layer_min(sample_summary), ".\n\n", sep = "")
cat("Spatial cropping min motif counts per layer: ",
    format_layer_min(space_summary), ".\n\n", sep = "")
cat("Random cell downsampling min motif counts per layer: ",
    format_layer_min(random_summary), ".\n\n", sep = "")
```


# Notes

- If runtime is too long, set `max_cells_per_image` or set `max_images`.
- Ensure `group_col` reflects a meaningful biological contrast with at least two groups.
- `motif_edger()` fits the volume-based Chung-Lu-style model by default; use `top_edges()` for edge results and
  `top_motifs2()` for wedge/triangle results with edge-submotif statistics.
