---
title: "CellEdgeR technical note"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CellEdgeR technical note}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Brief Intro

CellEdgeR builds Delaunay graphs from spatial cell coordinates, counts
label-based motifs, and tests differential abundance with edgeR using
Chung-Lu-style volume offsets. The default and primary statistical model is
volume based.

CellEdgeR tests two result sets:

- `edges`: edge motifs extracted with `top_edges()`.
- `motifs2`: wedge and triangle motifs extracted with `top_motifs2()`.

Node counts are used internally for exposure and submotif values, but nodes are
not tested. Wedges and triangles are kept separate; there is no merged `TP_*`
triplet test.

# Quick Start

```{r quick-start}
library(CellEdgeR)

samples_list <- make_demo_samples(n_cells = 2000, seed = 1)
metadata_df <- data.frame(
  condition = rep(c("ctrl", "treated"), each = length(samples_list) / 2),
  row.names = names(samples_list)
)

cellgraph <- build_cell_graphs(samples_list, n_cores = 2)
cellgraph <- count_motifs_graphs(cellgraph, max_edge_len = 50, include_wedge = TRUE, n_cores = 2)
cellgraph <- motif_edger(
  cellgraph,
  sample_df = metadata_df,
  design_formula = "~ condition"
)

top_edges(cellgraph, coef = "conditiontreated")
top_motifs2(cellgraph, coef = "conditiontreated")
```

# Motif Layers

For each sample, CellEdgeR builds a Delaunay triangulation on the 2D coordinates
and treats unique edges as candidate adjacency. Optional pruning by
`max_edge_len` removes long edges before counting motifs.

Motifs are counted by label:

- Nodes: `N_<label>`, used internally but not tested by `motif_edger()`.
- Edges: `E_<label1>_<label2>`, unordered.
- Triangles: `T_<label1>_<label2>_<label3>`, unordered.
- Wedges: `W_<center>_<leaf1>_<leaf2>`, centered and optional.

Counts are per sample, yielding a motif-by-sample matrix for each layer.

# Volume Offsets

The volume approach uses label-specific degree volumes to represent how much of
the graph is available for each label in each sample. For a label, the volume is
the sum of degrees of cells with that label. Let `2m` be twice the number of
edges in the sample.

Offsets are log expected counts under a Chung-Lu-style graph with preserved
label volumes:

- Edges: `log(vol_a) + log(vol_b) - log(2m)`, with a `-log(2)` correction for same-label unordered pairs.
- Triangles: `log(vol_a) + log(vol_b) + log(vol_c) - 2 * log(2m)`, plus an unordered-label multiplicity correction.
- Wedges: center exposure from `sum(degree * (degree - 1))` for the center label, leaf-label volumes, and an equal-leaf correction.

The offsets use a small `offset_pseudo` to avoid `log(0)`.

# Differential Model

`motif_edger()` stacks edge, wedge, and triangle counts and fits a single edgeR
quasi-likelihood negative-binomial GLM with volume offsets. The design matrix
comes from `design_formula` and can include multiple covariates:

```{r edger}
cellgraph <- motif_edger(
  cellgraph,
  sample_df = sample_df,
  design_formula = "~ condition + batch"
)

edges_tbl <- top_edges(cellgraph, coef = "conditiontreated")
motifs2_tbl <- top_motifs2(cellgraph, coef = "conditiontreated")
```

The logFC is interpreted as a differential motif rate after correcting for the
volume exposure. The p-value is a model coefficient test, not a direct test of
raw counts or normalized-count medians.

Before fitting, CellEdgeR applies `edgeR::filterByExpr()` to the stacked
edge/wedge/triangle count matrix. This is the standard edgeR low-expression
filter and prevents motifs observed in too few samples, including one-sample
spikes, from entering the multiple-testing table. The filtering summary is
stored in `cellgraph$edger$filter`.

# Motif2 Submotif Columns

`top_motifs2()` returns wedges and triangles in one table. It also adds the edge
submotifs implied by each three-label motif:

- `edge12`, `edge13`, `edge23`.
- `edge12_logFC`, `edge13_logFC`, `edge23_logFC`.
- `edge12_PValue`, `edge13_PValue`, `edge23_PValue`.
- `edge12_FDR`, `edge13_FDR`, `edge23_FDR`.
- `submotif_min_FDR` and `submotif_max_abs_logFC`.

For wedges, label 1 is the center. `edge12` and `edge13` are the center-leaf
edges, while `edge23` is the leaf-leaf closing edge.

# Required Inputs

1. `cells_by_sample`: a named list of data frames. Each data frame must have:
   - Column 1: numeric x coordinate.
   - Column 2: numeric y coordinate.
   - Column 3: label, character or factor.
2. `sample_df`: a data frame with one row per sample and row names equal to
   `names(cells_by_sample)`.

# Build Graphs And Count Motifs

```{r build-and-count}
cellgraph <- build_cell_graphs(cells_by_sample, n_cores = 4)
cellgraph <- count_motifs_graphs(
  cellgraph,
  max_edge_len = 50,
  include_wedge = TRUE,
  n_cores = 4
)
```

# Choose A Delaunay Cutoff

Inspect the edge-length distribution and choose a cutoff that removes the long
tail while keeping local neighbors. The cutoff is in the same units as your
coordinates.

```{r delaunay-cutoff}
edge_lengths <- unlist(lapply(cellgraph$per_sample_graph, function(ps) ps$edge_len))

ggplot2::ggplot(data.frame(edge_len = edge_lengths), ggplot2::aes(edge_len)) +
  ggplot2::geom_density(fill = "#4477aa", alpha = 0.3, color = "#223355") +
  ggplot2::scale_x_log10("Edge length (log scale)") +
  ggplot2::ylab("Density") +
  ggplot2::ggtitle("Edge length distribution (density, log-x)") +
  ggplot2::theme_minimal()
```

# Motif Space Size

`motif_space_size()` summarizes the number of possible tested edge, triangle,
and wedge label motifs:

```{r motif-space}
motif_space_size(cellgraph)
```

# Plotting Motifs

Spatial plot of a motif within a sample:

```{r plot-spatial}
plot_sample_graph(
  cellgraph,
  sample_name = "s1",
  max_edge_len = 50,
  motif_key = "T_a_b_c",
  motif_node_size = 0.8
)
```

Quantification of motifs across samples:

```{r plot-quantification}
norm_df <- get_motif_values(cellgraph, value = "norm")
df_plot <- data.frame(sample = rownames(norm_df), value = norm_df[["T_a_b_c"]])
df_plot <- merge(df_plot, sample_df, by.x = "sample", by.y = "row.names", all.x = TRUE)

ggplot2::ggplot(df_plot, ggplot2::aes(x = condition, y = value, fill = condition)) +
  ggplot2::geom_boxplot() +
  ggplot2::geom_jitter(width = 0.1, alpha = 0.6) +
  ggplot2::ylab("Normalized motif count (volume offset)") +
  ggplot2::xlab("condition")
```

# Assumptions

- Delaunay edges are a reasonable proxy for spatial adjacency in each sample.
- Pruning by `max_edge_len` removes long-range edges without removing meaningful local adjacency.
- Cell labels are accurate and consistent across samples.
- Samples are independent, and motif counts follow a negative-binomial model.
- Volume offsets approximate expected counts under a graph with preserved label degree volumes and total edge count.
